---
title: "Practical 1: Static model"
author: "Olivier Gimenez"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_depth: 2
date: "May 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      dpi = 300, 
                      fig.height = 6, 
                      fig.width = 1.777777*6)
```

# Objectives

+ Fit static (single-season) occupancy models
+ Deal with covariates
+ Carry out model selection

# Pre-requisites

Load some packages:
```{r}
library(tidyverse) # manipulate and visulaize data
theme_set(theme_light(base_size = 16)) # set theme for visualisation and font size
library(unmarked) # occupancy modelling
```

# Data

We use data from *Mapping and explaining wolf recolonization in France using dynamic occupancy models and opportunistic data* by Louvrier et al. in Ecography available at <https://onlinelibrary.wiley.com/doi/abs/10.1111/ecog.02874>. We use the year 2013 for illustration. Occasions are months December, January, February and March. A 100km$^2$ grid was overlaid over France and detections and non-detections were collected. See paper for more details. 

## Read in data

Read in data:
```{r}
wolf_data <- read_csv("data/wolf_occupancy.csv")
```

Inspect data. The column names are self-explanatory. Note that covariates effort (number of observers in a given cell), altitude and forest were standardized. 
```{r}
wolf_data
```

There are > 3000 cells. 
```{r}
nrow(wolf_data)
```

## Format data

We pick the detections and non-detections, and format them for use in `unmarked`:
```{r}
y <- wolf_data[,2:5]
wolf <- unmarkedFrameOccu(y = y)
```

Inspect:
```{r}
head(wolf)
```

Summarise data:
```{r}
summary(wolf)
```

Visualize:
```{r}
plot(wolf)
```

# Model with constant parameters

Fit a model with constant parameters, detection first, then occupancy:
```{r}
fm <- occu(~ 1 ~ 1, data = wolf)
```

Inspect results:
```{r}
fm
```

Parameter estimates are on the logit scale. We need to back-transform them on their natural scale, ie between 0 and 1. There are several ways to do just that. 

First, inspect model coefficients:
```{r}
coef(fm)
```

Get occupancy on logit scale, then back-transform it:
```{r}
logit_occupancy <- coef(fm)[1]
occupancy <- 1 / (1 + exp(-logit_occupancy))
occupancy
```

Idem for detection, note that we use the inverse logit function native to R:
```{r}
logit_det <- coef(fm)[2]
detection <- plogis(logit_det)
detection
```

Alternatively, you may use the `unmarked` function `backTransform`:
```{r}
backTransform(fm, type ='state')
backTransform(fm, type ='det')
```

And to get confidence intervals:
```{r}
confint(backTransform(fm, type='state'))
confint(backTransform(fm, type='det'))
```

# Deal with covariates

## Site-level covariates

Here we consider site-level covariates, which do not change between visits. 

Let's fit another model with detection as a function of effort, and occupancy as a function of forest cover. We need to recreate a dataset that includes these covariates, with one column per covariate:
```{r}
site.covs <- data.frame(effort = wolf_data$effort,
                        forest = wolf_data$foret)
wolf <- unmarkedFrameOccu(y = y,
                          siteCovs = site.covs)
```

Summarise:
```{r}
summary(wolf)
```

Fit the model:
```{r}
fm1 <- occu(~ effort ~ forest, data = wolf)
```

Inspect results:
```{r}
fm1
```

Now we'd like to plot the occupancy probabilty estimates as a function of forest cover, and detection probability estimates as a function of effort. 

First we create a grid of values for both covariates:
```{r}
effort_grid <- seq(min(site.covs$effort), 
             max(site.covs$effort), 
             length = 50)
forest_grid <- seq(min(site.covs$forest), 
             max(site.covs$forest), 
             length = 50)
```

Then we predict detection and occupancy probabilities using the parameter estimates:
```{r}
det_pred <- plogis(coef(fm1)[3] + coef(fm1)[4] * effort_grid)
occ_pred <- plogis(coef(fm1)[1] + coef(fm1)[2] * forest_grid)
```

Note that you can obtain the same result with function `predict`. E.g. let's consider detection:
```{r eval = FALSE}
effort <- effort_grid
predict(fm1, type = "det", as.data.frame(effort))
```

For occupancy, you would use:
```{r eval = FALSE}
forest <- forest_grid
predict(fm1, type = "state", as.data.frame(forest))
```

Now we can visualise both relationships:
```{r}
par(mfrow = c(1,2))
plot(forest_grid, 
     occ_pred, 
     type = "l", 
     lwd = 3, 
     xlab = "forest cover", 
     ylab = "estimated occupancy prob")
plot(effort_grid, 
     det_pred, 
     type = "l", 
     lwd = 3, 
     xlab = "effort", 
     ylab = "estimated detection prob")
```

## Observation-level covariates

We now consider for the observation-level covariates, which can change between visits. Each covariate is its own data.frame with rows as sites and columns as repeat visits.

Let's create a data.frame for a time covariate which we will use to assess the effect of temporal variation on detection:
```{r}
time <- matrix(c('visit 1','visit 2','visit 3', 'visit 4'),
              nrow = nrow(wolf@y),
              ncol = ncol(wolf@y), 
              byrow = TRUE)
```

Inspect:
```{r}
head(time)
```

We recreate our dataset with this new covariate:
```{r}
wolf <- unmarkedFrameOccu(y = y, # detection/non-detection data
                          siteCovs = site.covs, # site-level cov (effort and forest)
                          obsCovs = list(time = time)) # obs-level cov (time)
```

Summarize:
```{r}
summary(wolf)
```

Fit the model with an additive effect of effort and time on detection, and the effect of forest cover on occupancy:
```{r}
fm2 <- occu(~ effort + time ~ forest, wolf)
```

Inspect results:
```{r}
fm2
```

The parameter estimates of the relationship between occupancy probability and forest cover are similar to those obtained with a model without the time effect on detection. 

Now let's plot the effect of effort and time on detection. To do so, we may represent detection as a function of time for an average effort (which is 0 here because effort is standardised), or detection as a function of effort for the 4 visits.

Let us focus on detection as a function of effort for the 4 visits. First we create a grid of values for effort, and consider all combinations of values of effort and visits (1, 2, 3 or 4):
```{r}
effort_grid <- seq(min(site.covs$effort), 
             max(site.covs$effort), 
             length = 50)
time <- c('visit 1', 'visit 2', 'visit 3', 'visit 4')
grid <- expand_grid(effort = effort_grid, time = time)
```

Get estimates of detection probability:
```{r}
det_pred <- predict(fm2, type = "det", as.data.frame(grid))
```

Get vector or predicted values:
```{r}
pred <- det_pred$Predicted
```

Visualize:
```{r}
plot(effort_grid, 
     pred[grid$time == "visit 1"],
     type = "l", 
     lwd = 3, 
     xlab = "effort", 
     ylab = "estimated probability of detection",
     ylim = c(0,1))
lines(effort_grid, 
      pred[grid$time == "visit 2"],
      lwd = 3,
      col = "blue")
lines(effort_grid, 
      pred[grid$time == "visit 3"],
      lwd = 3,
      col = "pink")
lines(effort_grid, 
      pred[grid$time == "visit 4"],
      lwd = 3,
      col = "green")
legend("topleft", 
       col = c("black", "blue", "pink", "green"), 
       lwd = 3, 
       legend = c("december", "january", "february", "march"))
```


# Model selection

We fitted 3 models, and we'd like to ask the question, which one is the best supported by the data. To do so, we use the AIC. 

First we re-run our 3 models using the same dataset (`unmarked` will throw us an error message otherwise): 
```{r}
fm <- occu(~ 1 ~ 1, wolf)
fm1 <- occu(~ effort ~ forest, wolf)
fm2 <- occu(~ effort + time ~ forest, wolf)
```

Second, we gather our 3 models in a list:
```{r}
fmList <- fitList('{psi, p}' = fm, 
                  '{psi(forest), p(effort)}' = fm1, 
                  '{psi(forest), p(effort + time)}' = fm2)
```

Last, we get a table with AIC values for each model:
```{r}
modSel(fmList)
```

