---
title: "Practical 4: Multispecies models"
author: "Olivier Gimenez (adapted from Marco Salvatori's code)"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_depth: 2
date: "May 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      dpi = 300, 
                      fig.height = 6, 
                      fig.width = 1.777777*6)
```

# Objective

+ Fit occupancy models with co-occurrences between 2 or more species

# Pre-requisites

Load some packages:
```{r}
library(tidyverse) # manipulate and visualize data
theme_set(theme_light(base_size = 16)) # set theme for visualisation and font size
library(unmarked) # occupancy modelling
```

# Data

## Description

We use data from the paper *Co-occurrence of snow leopard, wolf and Siberian ibex under livestock encroachment into protected areas across the Mongolian Altai* by Salvatori et al. published in Biological Conservation and available at <https://www.sciencedirect.com/science/article/abs/pii/S0006320721003463>. The authors used camera-traps to sample four areas with different protection status across the Mongolian Altai Mountains, and targeted a predator-prey system composed of livestock, one large herbivore, the
Siberian ibex, and two large carnivores, the snow leopard and the wolf. Here we will focus on livestock and wolf. They showed attractino between livestock and wolf, with the potential for human-wildlife conflicts.

We thank Marco Salvatori for sharing the data.

The authors used a two-step approach to infer co-occurrence between species. In the first step they selected the most supported covariates for each species using single-species occupancy models. They first tested the distance to the closest herders' settlement and the camera performance as covariates on detection probability. They then tested the distance to the closest settlement, terrain slope, elevation, and the sampling area as categorical variable, on
occurrence probability while keeping the best encounter structure previously selected. In the second step, the most supported covariates from single-species occupancy models were included as predictors of occurrence and detectability of single species in the multi-species model. 

## Read in data

Read in covariates data:
```{r}
cov <- read.csv2('data/covs_original_ALL.csv', stringsAsFactors = F)
cov <- cov[,c('Sampling.Unit', 'Area', 'CT.sens', 'EL', 'SL', 'Dis')]
```

We have the site id (Sampling.Unit), area, camera performance (CT.sens), elevation (EL), terrain slope (SL) and distance to closest settlement (Dis). 

Standardize using `decostand()` function in `vegan` package:
```{r}
cov[,4:6] <- apply(X = cov[,4:6], MARGIN = 2, FUN = vegan::decostand, method = 'standardize')
```

Read in detections and non-detections:
```{r}
y <- readRDS(file = 'data/matrici_complessive.rds')
```

Select data for livestock and wolf
```{r}
y_livestock <- y$domestici
y_wolf <- y$lupo
```

# Single species analyses

## Organise data

```{r}
df_livestock <- unmarkedFrameOccu(y = y_livestock, siteCovs = cov)
df_wolf <- unmarkedFrameOccu(y = y_wolf, siteCovs = cov)
```

## Livestock covariates selection

Start with detection:
```{r}
mod0 <- occu(~1~Area+EL+SL+Dis, df_livestock)
mod1 <- occu(~Dis~Area+EL+SL+Dis, df_livestock)
mod2 <- occu(~CT.sens~Area+EL+SL+Dis, df_livestock)
mod3 <- occu(~Dis+CT.sens~Area+EL+SL+Dis, df_livestock)
```

Gather models and figure out best one:
```{r}
mods <- fitList('Zero' = mod0,
                'Dis' = mod1, 
                'CT.sens' = mod2, 
                'Dis + CT.sens' = mod3)
modSel(mods)
```

Proceed with occupancy
```{r}
mod1 <- occu(~Dis+CT.sens~1, df_livestock)
mod2 <- occu(~Dis+CT.sens~Area, df_livestock)
mod3 <- occu(~Dis+CT.sens~Dis, df_livestock)
mod4 <- occu(~Dis+CT.sens~EL, df_livestock)
mod5 <- occu(~Dis+CT.sens~SL, df_livestock)
mod6 <- occu(~Dis+CT.sens~Area+SL, df_livestock)
mod7 <- occu(~Dis+CT.sens~Area+EL, df_livestock)
mod8 <- occu(~Dis+CT.sens~Area+Dis, df_livestock)
mod9 <- occu(~Dis+CT.sens~EL+Dis, df_livestock)
mod10 <- occu(~Dis+CT.sens~EL+SL, df_livestock)
mod11 <- occu(~Dis+CT.sens~Dis+SL, df_livestock)
mod12 <- occu(~Dis+CT.sens~Area+EL+SL, df_livestock)
mod13 <- occu(~Dis+CT.sens~Area+EL+Dis, df_livestock)
mod14 <- occu(~Dis+CT.sens~Area+SL+Dis, df_livestock)
mod15 <- occu(~Dis+CT.sens~SL+EL+Dis, df_livestock)
mod16 <- occu(~Dis+CT.sens~ Area+EL+SL+Dis, df_livestock)
```

Gather models and figure out best one:
```{r}
mdls<-fitList('Zero'=mod0,
              'K'=mod1,
              'Area'=mod2,
              'Dis'=mod3,
              'EL'=mod4,
              'SL'=mod5,
              'Area+SL'=mod6,
              'Area+EL'=mod7, 
              'Area+Dis'=mod8, 
              'EL+Dis'=mod9, 
              'EL+SL'=mod10, 
              'Dis+SL'=mod11,
              'Area+EL+SL'=mod12, 
              'Area+EL+Dis'=mod13,  
              'Area+SL+Dis'=mod14, 
              'SL+EL+Dis'=mod15, 
              'Area+EL+SL+Dis'=mod16)
ms <- modSel(mdls)
```

Inspect best model:
```{r}
summary(mod13)
```

## Wolf covariates selection

Start with detection:
```{r}
mod0 <- occu(~1~Area+EL+SL+Dis, df_wolf)
mod1 <- occu(~Dis~Area+EL+SL+Dis, df_wolf)
mod2 <- occu(~CT.sens~Area+EL+SL+Dis, df_wolf)
mod3 <- occu(~Dis+CT.sens~Area+EL+SL+Dis, df_wolf)
```

Gather models and determine best one:
```{r}
mods <- fitList('Zero'=mod0,
                'Dis'=mod1,
                'CT.sens'=mod2, 
                'Dis+CT.sens'=mod3)
modSel(mods)
```

Proceed with occupancy:
```{r}
## psi
mod1 <- occu(~1~1, df_wolf)
mod2 <- occu(~1~Area, df_wolf)
mod3 <- occu(~1~Dis, df_wolf)
mod4 <- occu(~1~EL, df_wolf)
mod5 <- occu(~1~SL, df_wolf)
mod6 <- occu(~1~Area+SL, df_wolf)
mod7 <- occu(~1~Area+EL, df_wolf)
mod8 <- occu(~1~Area+Dis, df_wolf)
mod9 <- occu(~1~EL+Dis, df_wolf)
mod10 <- occu(~1~EL+SL, df_wolf)
mod11 <- occu(~1~Dis+SL, df_wolf)
mod12 <- occu(~1~Area+EL+SL, df_wolf)
mod13 <- occu(~1~Area+EL+Dis, df_wolf)
mod14 <- occu(~1~Area+SL+Dis, df_wolf)
mod15 <- occu(~1~SL+EL+Dis, df_wolf)
mod16 <- occu(~1~ Area+EL+SL+Dis, df_wolf)
```

Gather models, and determine best one:
```{r}
mdls <- fitList('Zero'=mod0,
              'K'=mod1,
              'Area'=mod2,
              'Dis'=mod3,
              'EL'=mod4,
              'SL'=mod5,
              'Area+SL'=mod6,
              'Area+EL'=mod7, 
              'Area+Dis'=mod8, 
              'EL+Dis'=mod9, 
              'EL+SL'=mod10, 
              'Dis+SL'=mod11,
              'Area+EL+SL'=mod12, 
              'Area+EL+Dis'=mod13,  
              'Area+SL+Dis'=mod14, 
              'SL+EL+Dis'=mod15, 
              'Area+EL+SL+Dis'=mod16)
modSel(mdls)
```

Inspect best model:
```{r}
summary(mod9)
```

## Visualisation

### Wolf

Fit best model again:
```{r}
modW <- occu(~1~EL+Dis, df_wolf)
```

Occupancy vs. distance to settlements:
```{r}
nd <- data.frame('Area' = factor(rep('KS', 216), levels = c('KS', 'SB', 'SU', 'TB')), 
                 'EL' = 0, 
                 'Dis' = cov$Dis, 
                 'SL' = 0)
wo <- predict(modW, type = 'state', newdata = nd)
wo$Distance <- cov$Dis

ggplot(data = wo, aes(x = Distance, y = Predicted)) +
  geom_line() + 
  geom_ribbon(aes(ymin = lower, ymax = upper ), alpha = .15) +
  ylim(0,1) + 
  labs(y = 'Estimated occupancy', x = 'Distance from settlements', title = 'Wolf') +
  scale_x_continuous(breaks = seq(min(wo$Dis), max(wo$Dis), length.out = 3),
                     labels = c("5000", "10000", "15000"))
```

Occupancy vs. elevation:
```{r}
nd <- data.frame('Area' = factor(rep('KS', 216), levels = c('KS', 'SB', 'SU', 'TB')), 
                 'Dis' = 0, 
                 'EL' = cov$EL, 
                 'SL' = 0)
wl <- predict(modW, type='state', newdata = nd)
wl$EL <- cov$EL

ggplot(data = wl, aes(x = EL, y = Predicted)) +
  geom_line() + 
  geom_ribbon(aes(ymin = lower, ymax = upper ), alpha = .15) +
  ylim(0,1) + 
  labs(y = 'Estimated occupancy', x = 'Elevation', title = 'Wolf') +
  scale_x_continuous(breaks=seq(min(wl$EL),max(wl$EL), length.out = 5),
                     labels=c("2000", "2400", "2800","3200", "3500"))
```

### Livestock

Fit best model again:
```{r}
modLIV <- occu(~Dis+CT.sens~Area+EL+Dis, df_livestock)
```

Occupancy vs. area:
```{r}
nd <- data.frame('Area' = factor(c('SB', 'TB', 'KS', 'SU'), levels =c('KS', 'SB', 'SU', 'TB') ), 
                 'EL' = rep(0,4), 
                 'SL' = rep(0,4), 
                 'Dis' = rep(0,4))
liv <- predict(modLIV, type = 'state', newdata = nd)
liv$Area<- factor(c('SB', 'TB', 'KS', 'SU'), levels=c('SB', 'TB', 'KS', 'SU'))

ggplot(data = liv, aes(x = Area, y = Predicted)) +
  geom_point(aes(x = Area, y = Predicted)) + 
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.5) +
  ylim(0,1) +
  labs(y = 'Estimated occupancy', title = 'Livestock')
```

Occupancy vs. elevation:
```{r}
nd <- data.frame('Area' = factor(rep('KS', 216), levels = c('KS', 'SB', 'SU', 'TB')), 
                 'EL' = cov$EL, 
                 'Dis' = 0, 
                 'SL' = 0)
liv <- predict(modLIV, type = 'state', newdata = nd)
liv$Elevation <- cov$EL

ggplot(data = liv, aes(x = Elevation, y = Predicted)) + 
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper ), alpha = .15) + 
  ylim(0,1) + 
  labs(y = 'Estimated occupancy', x = 'Elevation', title = 'Livestock') +
  scale_x_continuous(breaks = c(-2.1,-0.95, 0.3,1.56, 2.5),
                     labels = c("2000", "2400", "2800", '3200', '3500'))
```

Occupancy vs. distance to settlements:
```{r}
nd <- data.frame('Area' = factor(rep('KS', 216), levels = c('KS', 'SB', 'SU', 'TB')), 
                 'EL' = 0, 
                 'Dis' = cov$Dis, 
                 'SL' = 0) 
liv2 <- predict(modLIV, type='state', newdata = nd)
liv2$Distance <- cov$Dis

ggplot(data = liv2, aes(x = Distance, y = Predicted)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lower, ymax = upper ), alpha = .15) +
  ylim(0,1) + 
  labs(y = 'Estimated occupancy', x = 'Distance from settlements', title = 'Livestock') +
  scale_x_continuous(breaks = seq(min(liv2$Distance), max(liv2$Distance), length.out = 3),
                     labels = c("5000", "10000", "15000"))
```

# Co-occurrence between wolf and livestock

## Prep data

Put together the detection/non-detections for livestock and wolf, along with covariates:
```{r}
twosp_df <- unmarkedFrameOccuMulti(y = list('livestock' = y_livestock, 
                                       'wolf' = y_wolf), 
                              siteCovs = cov)
```

Summarize data:
```{r}
summary(twosp_df)
```

Visualize design:
```{r}
plot(twosp_df)
```

## Model with co-occurrence and constant parameters

We start by fitting a model with constant parameters:
```{r}
m0 <- occuMulti(detformulas = c('~1','~1'), # detection for livestock, then wolf
                stateformulas = c('~1', '~1', '~1'), # marginal occupancy for livestock, wolf, and co-occurrence
                data = twosp_df) # data
summary(m0)
```

Get estimated marginal occupancy for wolf:
```{r}
head(predict(m0, 'state', species = 'wolf'))
```

Marginal occupancy for livestock:
```{r}
head(predict(m0, 'state', species = 'livestock'))
```

Now the estimated probability of co-occurrence of wolf and livestock:
```{r}
head(predict(m0, 'state', species = c('wolf', 'livestock')))
```

What about wolf occupancy conditional on livestock being present:
```{r}
head(predict(m0, 'state', species = 'wolf', cond = 'livestock'))
```

And wolf occupancy conditional on livestock being absent:
```{r}
head(predict(m0, 'state', species = 'wolf', cond = '-livestock'))
```

Clearly, wolf is attracted by livestock.

A way to formally assess the existence of co-occurrence is to compare the model with co-occurrence to the model without co-occurrence. We fit the model without co-occurrence:

```{r}
m1 <- occuMulti(detformulas = c('~1','~1'), # detection for livestock, then wolf
                stateformulas = c('~1', '~1', '~0'), # marginal occupancy for livestock, wolf, and independence
                data = twosp_df) # data
```

Now compare AIC:
```{r}
tog <- fitList(m0, m1)
modSel(tog)
```

The model with co-occurrence has a slightly lower AIC value, suggesting co-occurrence happens between wolf and livetock.

## Model with co-occurrence and covariates

We fit a model with covariates (see single-species analyses) and co-occurrence between wolf and livestock:
```{r}
mod_twosp <- occuMulti(detformulas = c('~Dis+CT.sens','~1'), 
                       stateformulas = c('~Area+EL+Dis', '~EL+Dis', '~1'), 
                       data = twosp_df)
summary(mod_twosp)
```

We also fit a model without co-occurrence and compare their respective AIC:
```{r}
mod_ind <- occuMulti(detformulas = c('~Dis+CT.sens','~1'), 
                         stateformulas = c('~Area+EL+Dis', '~EL+Dis', '~0'), # set interaction to 0
                         data = twosp_df)

tog <- fitList(mod_twosp, mod_ind)
modSel(tog)
```

The model with co-occurrence has a lower AIC value than the model without co-occurrence. 

Now let's have a look to the estimated wolf occupancy probability conditional on livestock being absent or present: 

```{r}
z01 <- predict(mod_twosp, type = 'state', species = 'wolf', cond = '-livestock', newdata = cov)
z01$speciesB <- 'Conditional on livestock being'
z01$cond <- 'Absent'

z02 <- predict(mod_twosp, type = 'state', species='wolf', cond = 'livestock', newdata = cov)
z02$speciesB <- 'Conditional on livestock being'
z02$cond <- 'Present'

zz2 <- rbind(z01, z02)

zz2 %>% ggplot() + 
  geom_boxplot(aes(x = cond, y = Predicted)) +
  labs(x = "Conditional on livestock being", y = "Estimated wolf occupancy")
```
